---
title: "BME580_Project"
output: html_document
date: "2023-02-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(psych) # Contains the winsor function and other helpful statistical tools
library(tidyverse) # Remember from last homework that dplyr is included in tidyverse
library(gridExtra)
library(corrplot)
library(patchwork)
library(factoextra)
library(tidyverse)
library(corrplot)
```


```{r}
data_clinical_patient = read.csv('../dataset/data_clinical_patient.txt', sep='\t', header = TRUE)
data_clinical_patient[0:100,]
summary(data_clinical_patient)
```
```{r}
data_clinical_patient1 = data_clinical_patient %>% drop_na(OS_STATUS)
data_clinical_patient1 = data_clinical_patient1 %>% drop_na(SUBTYPE)
data_clinical_patient1 <- data_clinical_patient1[!(data_clinical_patient1$SUBTYPE==""),]
data_clinical_patient1$SUBTYPE = factor(data_clinical_patient1$SUBTYPE)

barchart = ggplot(data_clinical_patient1, 
                     aes(x=SUBTYPE, fill = OS_STATUS)) +
  geom_bar(position = "dodge") +
  labs(title = 'Overall Survival (OS) in Different Breast Cancer Subtypes',
       x = 'Breast Cancer Subtypes')
barchart


```

```{r}
data_mrna_seq_v2_rsem_zscores_ref_all_samples = read.csv('../dataset/data_mrna_seq_v2_rsem_zscores_ref_all_samples.txt', sep=' ', header = TRUE)
sprintf("number of rows = %s", nrow(data_mrna_seq_v2_rsem_zscores_ref_all_samples))
sprintf("number of cols = %s", ncol(data_mrna_seq_v2_rsem_zscores_ref_all_samples))
data_mrna_seq_v2_rsem_zscores_ref_all_samples[0:10,]
summary(data_mrna_seq_v2_rsem_zscores_ref_all_samples)
```
```{r}
mrna = data_mrna_seq_v2_rsem_zscores_ref_all_samples[c('X100130426', 'X100133144', 'UBE2Q2P2', 'HMGB1P1', 'X155060', 'X10431', 'RNU12.2P', 'SSX9P', 'X317712', 'EZHIP', 'EFCAB8', 'SRP14P1', 'X391343', 'TRIM75P', 'SPATA31B1P')]
mrna <- mrna[complete.cases(mrna),] 
mrna
pr_unscale = prcomp(mrna)
pr_scale = prcomp(mrna, scale=TRUE)
```

```{r}
names(pr_scale)
```

```{r}
dim(pr_scale$x)

pr_scale$x
```

```{r}
pr_scale$rotation
```
```{r}
fviz_eig(pr_scale, addlabels = T, ncp = 16)
fviz_eig(pr_unscale, addlabels = T, ncp = 16) # Why does the first PC of our unscaled data describe so much more variance than in the scaled data? Features with larger values appear to contribute more to the variance within the data compared to the smaller valued features when unscaled.

# Our loadings tell us that features with larger measurement (such as baseline.value) scales tend to contribute more to the PCs than features with smaller values ranges
pr_unscale$rotation
```
We can also looks at these in tabular form
```{r}
get_eig(pr_scale)
```


We can also visualize our first two PCs (which explain 54% of our cumulative variance) using a biplot. If you wished to visualize the first 3 principal components, you could create a 3D visualization with the plotly package.

To create our biplot, we'll use the factoextra package

```{r}
fviz_pca_biplot(pr_scale)
# We can observe some clusters, for example, the observations on the far left,
# and a cluster between those and the main set of observations in the middle
```
If we know our class labels, we can actually take this visualization one step further 
```{r}
fviz_pca_biplot(pr_scale, habillage = data$fetal_health)
```
We can also visualize this data without the PCs
```{r}
# Easier data visualization using factoextra
fviz_pca_ind(pr_scale,
             col.ind = 'coord',
             habillage = data$fetal_health)

```




```{r}
data_clinical_patient2 <- data_clinical_patient1[data_clinical_patient1$OS_STATUS=="1:DECEASED",]


barchart2 = ggplot(data_clinical_patient2, 
                     aes(x=SUBTYPE)) +
  geom_bar(position = "dodge") +
  labs(title = 'Mortality in Different Breast Cancer Subtypes',
       x = 'Breast Cancer Subtypes')
barchart2

```


```{r}
data_clinical_sample = read.csv('../dataset/data_clinical_sample.txt', sep='\t', header = TRUE)
data_clinical_sample[0:10,]
summary(data_clinical_sample)
```
```{r}
data_clinical_supp_hypoxia = read.csv('../dataset/data_clinical_supp_hypoxia.txt', sep='\t', header = TRUE)
data_clinical_supp_hypoxia[0:10,]
summary(data_clinical_supp_hypoxia)
```
```{r}
data_gene_panel_matrix = read.csv('../dataset/data_gene_panel_matrix.txt', sep='\t', header = TRUE)
data_gene_panel_matrix[0:10,]
summary(data_gene_panel_matrix)
```
```{r}
# takes forever to load
data_methylation_hm27_hm450_merged = read.csv('../dataset/data_methylation_hm27_hm450_merged.txt', sep='\t', header = TRUE)
data_methylation_hm27_hm450_merged[0:10,]
summary(data_methylation_hm27_hm450_merged)
```


```{r}
data_mutations = read.csv('../dataset/data_mutations.txt', sep='\t', header = TRUE)
data_mutations[0:10,]
summary(data_mutations)
```
```{r}
data_sv = read.csv('../dataset/data_sv.txt', sep='\t', header = TRUE)
data_sv[0:10,]
summary(data_sv)
#PROBMEL WITH COLUMNS
```
```{r}
data_timeline_sample_acquisition = read.csv('../dataset/data_timeline_sample_acquisition.txt', sep='\t', header = TRUE)
data_timeline_sample_acquisition[0:10,]
summary(data_timeline_sample_acquisition)
```
```{r}
data_timeline_status = read.csv('../dataset/data_timeline_status.txt', sep='\t', header = TRUE)
data_timeline_status[0:10,]
summary(data_timeline_status)
```
```{r}
data_timeline_treatment = read.csv('../dataset/data_timeline_treatment.txt', sep='\t', header = TRUE)
data_timeline_treatment[0:10,]
summary(data_timeline_treatment)
```
